### 2019-09-28 paper-two/paper-two.tex
switching explained, dmn started

\begin{center}
  \Huge{Context}
\end{center}

I took my first Vipassana course in 2017. Vipassana meditation was so unlike anything
I had ever experienced before, after the course I was left extremely curious
about what exactly had happened to me during those ten days. For months afterward, I
spent my mornings and evenings wading through the research in the field of
meditation, with a specific focus on white papers dealing with the effects a
10-day Vipassana course has on the brains of participants. The research in this area
is limited. The quality research which does exist usually uses a sample of highly
experienced meditators rather than beginners and/or self-reports rather than
objective measures.

@@

This paper does not contradict Mr. Goenka's sentiment, but instead acts as a starting
point for those who view the brain as the quintessential physical component of mind.

\textbf{Disclaimer:} Although my primary field of research is biology, my work is far removed
from neuroscience. I have tried to simplify the research available so as to better
understand it. If there are any corrections to make or editing in terms of the
content, please feel free to get in touch with me.

\begin{center}
  \Huge{Brain Function and Anatomy}
\end{center}

Before dissecting the experience of meditation as it pertains to the brain, this
section describes the different parts of the brain that have shown up in
scientific literature as correlated to meditative practices. Neuroscience divides the
brain into the reified geography of the brain, its anatomy, and the abstract concepts
governed by brain activity, its function. Between concrete physiology and abstract
outcomes exist networks of cooperative structures which correspond to general
high-level activities of the brain.

Default Mode Network:
These are a set of structures in the mid-line of the brain (which is a evolutionarily more primitive area related to memory and emotion) which connects to structures in the cortex (which is evolutionarily the most recent part of the brain and contains the executive and higher order functions). It is the constellation of regions that light up when the brain is not engaged in any external/goal-oriented tasks (1). These regions are functionally connected and are involved in
a) The neurological basis of the self —
The regions involved in this are the Posterior Cingulate Cortex (PCC), venteromedial Prefrontal Cortex (vmPFC) and the Inferiour Parietal Lobule (IPL) (1).
Cerebral blood flow and metabolic rate is 40\% higher in the PCC compared to the average across the brain making it one of the most active regions of the brain. (Leech R, Sharp DJ (July 2013). “The role of the posterior cingulate cortex in cognition and disease”. Brain. 137 (Pt 1): 12–32.)
The PCC has been thought to mediate interactions between emotion and memory. It consistently lights up when a person recollects something from their life and depending on the emotional importance of the memory the strength of the lighting up changes. It is also activated by emotional stimuli both positive and negative and acts as an interface between the world and us by gauging the importance of different stimuli(2).

@@

Andrews-Hanna, Jessica R. (2012–06–01). “The brain’s default network and its adaptive role in internal mentation”. The Neuroscientist: A Review Journal Bringing Neurobiology, Neurology and Psychiatry. 18 (3): 251–270. doi:10.1177/1073858411403316. ISSN 1089–4098. PMC 3553600. PMID 21677128.
Maddock, Richard J.; Garrett, Amy S.; Buonocore, Michael H. (January 2003). “Posterior cingulate cortex activation by emotional words: fMRI evidence from a valence decision task”. Human Brain Mapping. 18 (1): 30–41.
Andrews-Hanna, Jessica R.; Smallwood, Jonathan; Spreng, R. Nathan (2014–05–01). “The default network and self-generated thought: component processes, dynamic control, and clinical relevance”. Annals of the New York Academy of Sciences. 1316(1): 29–52. doi:10.1111/nyas.12360. ISSN 1749–6632. PMC 4039623. PMID 24502540.
Davey CG, Pujol J, Harrison BJ. Mapping the self in the brain’s default mode network. Neuroimage. 2016 May 15;132:390–397.
Isoda, M., \& Noritake, A. (2015). What makes the dorsomedial frontal cortex active during reading the mental states of others?. Neural basis of social learning, social deciding, and other-regarding preferences, 51.
Waytz, A., Zaki, J., \& Mitchell, J. P. (2012). Response of dorsomedial prefrontal cortex predicts altruistic behavior. The Journal of Neuroscience, 32(22), 7646–7650.
Menon, V; Toga, A (2015). Salience Network. Elsevier. pp. 597–611. ISBN 978–0–12–397316–0.
Pardo JV, Pardo PJ, Janer KW, Raichle ME (January 1990). “The anterior cingulate cortex mediates processing selection in the Stroop attentional conflict paradigm”. Proceedings of the National Academy of Sciences of the United States of America. 87 (1): 256–9. doi:10.1073/pnas.87.1.256. PMID 2296583.
Bush G, Vogt BA, Holmes J, Dale AM, Greve D, Jenike MA, Rosen BR (January 2002). “Dorsal anterior cingulate cortex: a role in reward-based decision making”. Proceedings of the National Academy of Sciences of the United States of America. 99 (1): 523–8. doi:10.1073/pnas.012470999. PMC 117593. PMID 11756669.
Sevinc G, Gurvit H, Spreng RN (July 2017). “Salience network engagement with the detection of morally laden information”. Social Cognitive and Affective Neuroscience. 12 (7): 1118–1127. doi:10.1093/scan/nsx035. PMID 28338944.
Jackson PL, Brunet E, Meltzoff AN, Decety J (2006). “Empathy examined through the neural mechanisms involved in imagining how I feel versus how you feel pain”. Neuropsychologia. 44 (5): 752–61. doi:10.1016/j.neuropsychologia.2005.07.015. PMID 16140345.
Bush G, Luu P, Posner MI (June 2000). “Cognitive and emotional influences in anterior cingulate cortex”. Trends in Cognitive Sciences. 4 (6): 215–222. doi:10.1016/S1364–6613(00)01483–2. PMID 10827444
Gianaros PJ, Derbyshire SW, May JC, Siegle GJ, Gamalo MA, Jennings JR. Anterior cingulate activity correlates with blood pressure during stress. Psychophysiology. 2005;42(6):627–35.
Xiaosi Gu, Patrick R. Hof, Karl J. Friston, Jin Fan. J Comp Neurol. 2013 Oct 15; 521(15): 3371–3388.
Rolls ET. Functions of the anterior insula in taste, autonomic, and related functions. Brain Cogn. 2016 Dec;110:4–19.
Lazar SW; Bush George; Gollub RL.; Fricchione, GL.; Khalsa G; Benson H. Functional brain mapping of the relaxation response and meditation.
Neuroreport. 11(7):1581–1585, May 15, 2000.
J. Muñoz-Ortiz, E. Muñoz-Ortiz, L. López-Meraz, L. Beltran-Parrazal, C. Morgado-Valle. The pre-Bötzinger complex: Generation and modulation of respiratory rhythm, Neurología (English Edition), 2018, ISSN 2173–5808,
https://doi.org/10.1016/j.nrleng.2018.05.006.
REMPEL‐CLOWER, N. L. (2007), Role of Orbitofrontal Cortex Connections in
Emotion. Annals of the New York Academy of Sciences, 1121:
72–86. doi:10.1196/annals.1401.026

@@

>>>>

\section{Introduction}

I took my first Vipassana course in 2017. Vipassana meditation was so unlike anything
I had ever experienced before, after the course I was left extremely curious
about what exactly had happened to me during those ten days. For months afterward, I
spent my mornings and evenings wading through the research in the field of
meditation. I was specifically focused on white papers dealing with the effects of a
10-day Vipassana course on the brains of participants. The research in this area
is limited. The quality research which does exist usually uses a sample of highly
experienced meditators rather than beginners and/or self-reports rather than
objective measures.

@@

This paper does not contradict Mr. Goenka's sentiment, but instead acts as a starting
point for readers who view the function of the brain as central to the activity of
the mind.

\textbf{Disclaimer:} Although my primary field of research is within the field of
biology, my work is far removed
from neuroscience. I have tried to simplify the research available so as to better
understand it. If there are any corrections to make or editing in terms of the
content, please feel free to get in touch with me.

\section{Brain Function and Anatomy}

Before dissecting the experience of meditation as it pertains to the brain, this
section describes the different parts of the brain that have shown up in
scientific literature as correlated to meditative practices. Recent neuroscience divides the
brain into the reified geography of the brain, its anatomy, and the abstract concepts
governed by brain activity, its function. Between concrete physiology and abstract
functional outcomes exist networks of cooperative structures which correspond to general
high-level activities of the brain.

The Default Mode Network (DMN) is the constellation of regions which fire when
the brain is not engaged in any external or goal-oriented
tasks. As the name suggests, the brain is in its ``default
mode''. \cite{defaultnetworkadaptive} The DMN is anticorrelated to the Central Executive
Network (CEN), which is active during externally-directed, high-level cognitive
functions. \cite{saliencenetwork} This anticorrelation between the DMN and the CEN is
governed by the Salience Network (SN), the collection of regions in the brain which
help decide which stimuli deserve our attention. The SN acts as a switch between the
internally-directed DMN and the externally-directed CEN. \cite{saliencenetwork}

[ TODO: insert switching diagram ]

\subsection{Default Mode Network}

The DMN is comprised of specific anatomy including portions of the mid-line of the
brain, an evolutionarily primitive area related to memory and emotion, and
structures in the cortex, an evolutionarily recent part of the brain containting the
executive and higher order functions. \cite{defaultnetworkanatomy}

These regions are functionally connected and are involved in the neurological basis
of the self, considering the mental states of others, remembering the past, and
imagining the future.

a) The neurological basis of the self —
The regions involved in this are the Posterior Cingulate Cortex (PCC), venteromedial Prefrontal Cortex (vmPFC) and the Inferiour Parietal Lobule (IPL) (1).
Cerebral blood flow and metabolic rate is 40\% higher in the PCC compared to the average across the brain making it one of the most active regions of the brain. (Leech R, Sharp DJ (July 2013). “The role of the posterior cingulate cortex in cognition and disease”. Brain. 137 (Pt 1): 12–32.)
The PCC has been thought to mediate interactions between emotion and memory. It consistently lights up when a person recollects something from their life and depending on the emotional importance of the memory the strength of the lighting up changes. It is also activated by emotional stimuli both positive and negative and acts as an interface between the world and us by gauging the importance of different stimuli(2).

@@

Maddock, Richard J.; Garrett, Amy S.; Buonocore, Michael H. (January 2003). “Posterior cingulate cortex activation by emotional words: fMRI evidence from a valence decision task”. Human Brain Mapping. 18 (1): 30–41.
Andrews-Hanna, Jessica R.; Smallwood, Jonathan; Spreng, R. Nathan (2014–05–01). “The default network and self-generated thought: component processes, dynamic control, and clinical relevance”. Annals of the New York Academy of Sciences. 1316(1): 29–52. doi:10.1111/nyas.12360. ISSN 1749–6632. PMC 4039623. PMID 24502540.
Davey CG, Pujol J, Harrison BJ. Mapping the self in the brain’s default mode network. Neuroimage. 2016 May 15;132:390–397.
Isoda, M., \& Noritake, A. (2015). What makes the dorsomedial frontal cortex active during reading the mental states of others?. Neural basis of social learning, social deciding, and other-regarding preferences, 51.
Waytz, A., Zaki, J., \& Mitchell, J. P. (2012). Response of dorsomedial prefrontal cortex predicts altruistic behavior. The Journal of Neuroscience, 32(22), 7646–7650.

Pardo JV, Pardo PJ, Janer KW, Raichle ME (January 1990). “The anterior cingulate cortex mediates processing selection in the Stroop attentional conflict paradigm”. Proceedings of the National Academy of Sciences of the United States of America. 87 (1): 256–9. doi:10.1073/pnas.87.1.256. PMID 2296583.
Bush G, Vogt BA, Holmes J, Dale AM, Greve D, Jenike MA, Rosen BR (January 2002). “Dorsal anterior cingulate cortex: a role in reward-based decision making”. Proceedings of the National Academy of Sciences of the United States of America. 99 (1): 523–8. doi:10.1073/pnas.012470999. PMC 117593. PMID 11756669.
Sevinc G, Gurvit H, Spreng RN (July 2017). “Salience network engagement with the detection of morally laden information”. Social Cognitive and Affective Neuroscience. 12 (7): 1118–1127. doi:10.1093/scan/nsx035. PMID 28338944.
Jackson PL, Brunet E, Meltzoff AN, Decety J (2006). “Empathy examined through the neural mechanisms involved in imagining how I feel versus how you feel pain”. Neuropsychologia. 44 (5): 752–61. doi:10.1016/j.neuropsychologia.2005.07.015. PMID 16140345.
Bush G, Luu P, Posner MI (June 2000). “Cognitive and emotional influences in anterior cingulate cortex”. Trends in Cognitive Sciences. 4 (6): 215–222. doi:10.1016/S1364–6613(00)01483–2. PMID 10827444
Gianaros PJ, Derbyshire SW, May JC, Siegle GJ, Gamalo MA, Jennings JR. Anterior cingulate activity correlates with blood pressure during stress. Psychophysiology. 2005;42(6):627–35.
Xiaosi Gu, Patrick R. Hof, Karl J. Friston, Jin Fan. J Comp Neurol. 2013 Oct 15; 521(15): 3371–3388.
Rolls ET. Functions of the anterior insula in taste, autonomic, and related functions. Brain Cogn. 2016 Dec;110:4–19.
Lazar SW; Bush George; Gollub RL.; Fricchione, GL.; Khalsa G; Benson H. Functional brain mapping of the relaxation response and meditation.
Neuroreport. 11(7):1581–1585, May 15, 2000.
J. Muñoz-Ortiz, E. Muñoz-Ortiz, L. López-Meraz, L. Beltran-Parrazal, C. Morgado-Valle. The pre-Bötzinger complex: Generation and modulation of respiratory rhythm, Neurología (English Edition), 2018, ISSN 2173–5808,
https://doi.org/10.1016/j.nrleng.2018.05.006.
REMPEL‐CLOWER, N. L. (2007), Role of Orbitofrontal Cortex Connections in
Emotion. Annals of the New York Academy of Sciences, 1121:
72–86. doi:10.1196/annals.1401.026

@@

\bibitem{defaultnetworkadaptive}
  Andrews-Hanna, Jessica R.
  \textit{The brain’s default network and its adaptive role in internal mentation.}
  The Neuroscientist: A Review Journal Bringing Neurobiology, Neurology and
  Psychiatry. 18 (3): 251–270. doi:10.1177/1073858411403316. ISSN 1089–4098. PMC
  3553600. PMID 21677128. (2012–06–01).

\bibitem{defaultnetworkanatomy}
  Buckner, R. L.; Andrews-Hanna, Jessica R.; Schacter, D. L.
  \textit{The Brain's Default Network: Anatomy, Function, and Relevance to Disease.}
  Annals of the New York Academy of Sciences. 1124 (1): 1–38. CiteSeerX
  10.1.1.689.6903. doi:10.1196/annals.1440.011. PMID 18400922, 2008

\bibitem{saliencenetwork}
  Menon, V; Toga, A.
  \textit{Salience Network.}
  Elsevier. pp. 597–611, 2015
  ISBN: 978–0–12–397316–0.
